---
title: "Machine Learning Course Project"
author: "Juan Sebastian Cuervo"
date: "15 de julio de 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

We perform the following data analysis and cleaning with the databases and using two models the Decision Tree and Random Forest we buil our Machine Learning Algorithms 

## Loading packages and data 

We initially load all required data and libraries in R studio 

```{r}
## Load required libraries into RStudio 

library(lattice)
library(ggplot2)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(corrplot)
library(rattle)


## Prepare Files in Working Directory taking Coursera URL 

urlTrain <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
urlTest <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

Trainarchivo <- "pml-training.csv"
TestArchivo <- "pml-testing.csv"

if (!file.exists(Trainarchivo)) 
{
  download.file(urlTrain, destfile=Trainarchivo, mode ="wb")
}
if (!file.exists(TestArchivo)) 
{
  download.file(urlTest, destfile=TestArchivo, mode ="wb")
}


## Load data in RSTudio. As there are blanks we assign them NA value 

DataTrain <- read.csv("pml-training.csv",na.strings = c("","NA"))
DataTest <- read.csv("pml-testing.csv")

```

## Data Cleaning 

There are a great number of columns with missing values so we take them out of database

```{r}
DataTrain<- DataTrain[,!sapply(DataTrain,function(DataTrain) any(is.na(DataTrain)))]
DataTest<- DataTest[,!sapply(DataTest,function(DataTest) any(is.na(DataTest)))]
```


## Data Partition into Test and Train Data sets

```{r}
set.seed(22519)
Cortar<- createDataPartition(DataTrain$classe, p=0.70, list=F)
DataTrainFinal <- DataTrain[Cortar, ]
DataValidation<- DataTrain[-Cortar, ]
```

## First Machine Learning Algorithm - Decision Tree 

We estimate the decision tree and calculate predictions and confusion matrix 

```{r}

## Model Fitting 1 - Decision Tree Model 

treeModel <- rpart(classe ~ ., data=DataTrainFinal, method="class")
fancyRpartPlot(treeModel)

## Prediction - Confusion Matrix  - Decision Tree  

predictionspower <- predict(treeModel, DataValidation , type = "class")
confusionMatrix(predictionspower, DataValidation$classe)

```

## Second Machine Learning Algorithm - Randome Forest 

We estimate the decision tree and calculate predictions and confusion matrix 

```{r}

## Model Fitting 2 - Random Forest

RanForModel<- randomForest(classe ~. , data=DataTrainFinal)

predictionpower2 <- predict(RanForModel, DataValidation, type = "class")
confusionMatrix(predictionpower2, DataValidation$classe)

## As we obtian a better fit,we calculate the out of sample error of this code

outofsample <- 1- as.numeric(confusionMatrix(DataValidation$classe,predictionpower2)$overall[1])
outofsample
```

## Bibliography

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises.
Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) .
Stuttgart, Germany: ACM SIGCHI, 2013

 
